---
title: "Session 1: R Basics"
author: "Ya-Feng Wen"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, results=FALSE)
```

# Objectives
+ Brief introduction of RStudio
+ Install packages
+ Basic date type and structures


# Download R and RStudio
+ R: <http://lib.stat.cmu.edu/R/CRAN/>
+ RStudio: <https://rstudio.com/>
+ Check R version: `sessionInfo()` or `version`
+ Update R: use `installr` package for Windows and `updateR` package for Mac. Detailed instruction can be found [here](https://www.linkedin.com/pulse/3-methods-update-r-rstudio-windows-mac-woratana-ngarmtrakulchol/).
    - One benefit of using `installr` is that it will update R and all its package
```{r, eval=F}
# installing/loading the package:
if(!require(installr)) {
install.packages("installr"); require(installr)} #load / install+load installr
 
# using the package:
updateR()
```

+ Update RStudio: `Help`>`Check for Updates`

# Install packages
+ Available packages on R CRAN: <https://cran.r-project.org/>
+ Developed and maintained by RStudio: <https://rstudio.com/products/rpackages/> 
+ Install stable release packages from CRAN: `install.packages("the package's name")` Only need to do this once. 
+ Use the package: `library(the package's name)`. Need for every session. 
+ Update packages: `update.packages(ask = FALSE)` (only update those on CRAN)
+ If packages is not avialble in CRAN, use `devtools::install_github("URL")`
+ Update all packages after updating R: 

```{r, eval=F}
## get packages installed
packs = as.data.frame(installed.packages(.libPaths()[1]), stringsAsFactors = F)

## and now re-install install packages using install.packages()
install.packages(packs$Package)
```

# Why use RStudio?
+ It's interactive. You can test your code (console) and document (script, markdown, slides, etc.) simultaneously.
+ It has four panels: Console, Scripts, Environments, Files/Plots/Packages/Help.
+ Publish your work to RPubs. [Example](https://rpubs.com/Statinmylife/473540)
+ Autocompletion to save time and reduce errors.
+ Excute code directly from RStudio script by using the <kbd>Ctrl</kbd> + <kbd>Enter</kbd> in Windows or <kdb>Return</kbd> in Mac.
+ <kbd>Ctrl</kbd> + <kbd>1</kbd> and <kbd>Ctrl</kbd> + <kbd>2</kbd> jump between the script and the console windows.
+ Use <kbd>#</kbd> signs to comment. (comment as detailed as you can)

# Basic Data Types and Data Structures in R
**Everything** in R is an object. The object can be character, integer, floating point,  Boolean etc. Unlike C and java, variables (reserved memory locations to store values) are not declared as some data type. The variables are assigned with R-Objects and the data type of the R-object becomes the data type of the variable.

Commonly used objects: 

+ **Vectors**: use `vector()` to create a vector. There are 6 data types of these atomic vectors. (When we call a vector atomic, we mean that the vector only holds data of a single data type.) A vector is one-dimensional and each element is of the same type.
+ **Lists**: `list()`. A list can contain many different types of elements, like vectors, functions or another list. It can be extremely useful inside functions. Since functions in R can only return a single object, list can "staple" many different results into a single object.
+ **Matrices**: `matrix()`. A matrix is a two-dimensional rectangular data set.
+ **Arrays**: `array()`. A array can be of any number of dimensions. 
+ **Factors**: `factor()`. Factors are the r-objects which are created using a vector. It stores the vector along with the distinct values of the elements in the vector as labels. 
+ **Data Frames**: `data.frame()`. Data frames are tabular data objects.

Six cleasses of atomic vectors:

+ **Logical**: `TRUE`, `FALSE`
+ **Numeric**: `2`, `2.5`
+ **Integer**: `2L`, `0L`
+ **Complex**: `3+2i`
+ **Character**:`"a"`, `"good"`, `"TRUE"`
+ **Raw**: `"Hello"` is stored as 48 65 6c 6c 6f (I've never used it.)

Functions to check the features of vectors and other objects: 

+ `class()`: what kind of object is it? (high level)
+ `typeof()`: what is the data type of this object? (low level)
+ `length()`: how long is it?
+ `attributes()`: does it have any metadat?

Functions to check the atrributes of objects:

+ `names()`
+ `dimnames()`
+ `dim()`

Functions for data frame (commonly used):

+ `head()` - shows first 6 rows
+ `tail()` - shows last 6 rows
+ `dim()` - returns the dimensions of data frame (i.e. number of rows and number of columns)
+ `nrow()` - number of rows
+ `ncol()` - number of columns
+ `str()` - structure of data frame - name, type and preview of data in each column
+ `names()` or `colnames()` - both show the names attribute for a data frame
+ `sapply(dataframe, class)` - shows the class of each column in the data frame

## Assignment Operator
+ Use `<-` for assignments. (Short cut: <kbd>Alt</kbd> + <kbd>-</kbd>)
+ Use `=` for specify the values of arguments in functions.

## Available data sets in R
Use `data()` to view all available dataset. Call the data set by its name directly. For example: `ChickWeight`,`iris`

## Practice
1. Create a variable called 'x' and store the number from 0 to 5. Check the features of x.
```{r}
x <- 0:5
# alternatively you can use seq(). Use `? function` to look up the function
x <- seq(5) # does not contain 0
x <- seq(0,5)
x
class(x); typeof(x); length(x); attributes(x); str(x)

# convert to numeric
y <- as.numeric(x)
y
class(y); typeof(y)

# convert to character
z <- as.character(x)
z
class(z); typeof(z)

# convert to logical
l <- as.logical(x)
l
class(l); typeof(l)
```

2. Create an empty numeric vector with 6 elements and store it as 'a'. Then update 'a' to 6 consequtive even numbers start from 2.
```{r}
a <- vector("numeric", length=6)
a <- numeric(6)

a <- c(2,4,6,8,10,12)
typeof(a); str(a)

a <- seq(from = 2, to = 12, by = 2) 
a <- seq(2,12,2) # alternative
length(a); str(a)
```

3. Guess the type of following vector: `b <- c(2.2, "a")`, `c <- c(FALSE, 1)`, `d <- c("a", FALSE)`
```{r}
b <- c(2.2, "a")
typeof(b)

c <- c(FALSE, 1)
typeof(c)

d <- c("a", FALSE)
typeof(d)

# R will convert all the elements to the same type. This is "coercion". When R converts the mode of storage based on its content, it is referred to as “implicit coercion”. 
# This may cause unwanted behavior.
```

4. Create a 2x2 identity matrix and store it to 'I'. Create another 2x3 matrix with first row of 1,3,5 and second row of 2,4,6 and store it to 'M'. What is the element of 'M' at 2,3 position?
```{r}
I <- matrix(
  c(1,0,0,1),
  nrow=2,
  ncol=2
)

I <- diag(2) # alternative way

M <- matrix(1:6, nrow=2, ncol=3)
# alternative way
M <- 1:6
dim(M) <- c(2,3)
M
# alternative way, using `rbinc()` and `cbind()`
x <- seq(1,5,2)
y <- seq(2,6,2)
rbind(x,y)

# Use `byrow` argument to specify how the matrix is filled
matrix(1:6, nrow=2, ncol=3, byrow=FALSE)

# Using single sqaure brackets to access to the element. It is always "row" then "column".
M[2.3]
```

5. Create a list called list1 containing 4 elements: 1) an integer vector of 1,2,3; 2) an character "Pharmacometrix"; 3) a function `sin`; 4) dataset: iris. What is the object type of each element?

```{r}
list1 <- list(c(1L,2L,3L), "Pharmacometrix", sin, iris)
list1
str(list1)

a <- list1[1]
# a + 3
class(list1[1])

# The content of elements of a list can be retrieved by using double square brackets
b <- list1[[1]]
b + 3
class(list1[[1]])

# Name each element is the list
names(list1) <- c("integer vector", "character", "functions", "dataset")
list1

# alternatively, you can add the name while creating the list
list2 <- list(`integer vector` = c(1L,2L,3L), 
              character = "Pharmacometrix", 
              functions = sin, 
              dataset = iris)
list2

# After adding a name for each element in the list, you can access by using `$` notation
chr1 <- list1$character
chr1
```

6. Create a dataframe for 10 subjects with the following columns and values: 
    + ID: 1 to 10
    + SEX: randomly select "male", "female", NA. Make this column a factor. 
    + AGE: follow a normal distribution with mean 60 and standard deviation of 10

What is the object type for each column?
How many missing values?

```{r}
set.seed(1) # to create a reproducible result
df <- data.frame(
  ID = 1:10,
  SEX = as.factor(sample(c("male", "female", NA), size=10, replace=TRUE)), # create SEX as a factor
  AGE = round(rnorm(10, mean = 60, sd = 10),0) # round the age to a whole number
)

# Check the object type of this dataframe
str(df); dim(df); nrow(df); ncol(df); names(df)

# Check if there is any missing values
anyNA(df)
# Count how many missing values
sum(as.numeric(is.na(df)))
# Using apply function, same as using a loop (will cover in the next session)
sapply(df, function(x) sum(is.na(x)))

apply(df, MARGIN = 2, function(x) sum(is.na(x)))
# count the NAs row-wise (case by case)
apply(df, MARGIN = 1, function(x) sum(is.na(x)))

```


# Resources: 
1. R Tutorial <https://www.tutorialspoint.com/r/index.htm>
2. R for Data Science by Grlemund and Wickham (2017), online at <https://r4ds.had.co.nz/>
3. R cheetsheets <https://rstudio.com/resources/cheatsheets/>
4. Carpentry Programming with R <https://swcarpentry.github.io/r-novice-inflammation/>

# Session Info:
```{r, results=TRUE}
sessionInfo()
```

